Left-linear rules in adhesive categories

### 2024-02-05 icalp/fusione.tex
minor changes to intro

Rewriting over many kind of graphs is elegantly captured by the
so-called \emph{double pushout} (DPO) approach to rewriting. A
noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their
variants~\cite{lack2005adhesive,ehrig2006weak}. Operating within the
framework of adhesive categories allows one to devise the foundations
of a rewriting theory which can be subsequently instantiated to the
context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting.

Rewriting over adhesive categories has been widely studied in the last
two decades and most of the research focused on the so-called linear
rewriting rules. Roughly speaking, linear rewriting rules can specify
that, in a rewriting step, some entities in the current state are
removed, some other are needed for the rewriting step to occur but
remain unaltered and some new entities can be generated.

@@

In some situations rewriting with linear rules are not sufficiently
expressive for capturing the dynamic of a system. This is the case
when a computational step is required not only to consume, preserve
and generated, but also ``merge'' parts of the state. This happens,
e.g., in nominal calculi where, as a result of name passing, the
received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimizations~\cite{WNW:egg}. Rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

@@

Technically, to acquire the expressiveness needed for expressing
fusions, i.e., the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories (which capture most
``applications'') have not been systematically explored. ADD REFS

@@

Firstly, Church-Rosser theorem does not work out of the box. In
particular, sequential independence between rewriting steps~\cite{}
expressed in terms of the existence of a so-called independence pair
is a canonical and well-understood notion of independence in the
linear case. \todo{P: Too technical? Forse qui e' piu' esplicativo un
  esempio} Whenever two steps are sequential independent
\begin{itemize}
\item they can be switched and 
\item the result of the switch is uniquely determined.
\end{itemize}

Here we observe that, in general, both property can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  A refined notion of independence can be considered to ensure
that independent rewriting steps can be switched, based in the concept
of a so-called filler. The problem was identified already
in~\cite{baldan2011adhesivity} where a suitable subclass of adhesive
categories is identified where independence as induced by a filler
coincide with sequential independence (as induced by an independence
pair). Here this is extended to $\mathcal{M}$-adhesive categories.

@@

Rather we focus on a class of rewriting systems, the so-called
\emph{very tame rewriting systems} in $\mathcal{M}$-adhesive
categories where sequential independence ensures existence and
uniqueness of the switch and argue that this is the right setting for
dealing with left-linear rewriting systems.

On the one hand we argue that very tame rewriting systems capture most
categories of interest for rewriting. In particular, we single out a
class of left-linear rewriting systems on graph-like structures which
are very tame and we show that rewriting systems on (term)graphs with
equivalences resempling e-graphs are also very tame. As a sanity
check, also linear rewriting systems on $\mathcal{M}$-adhesive
categories are very tame.

On the other hand we show that an appropriate theory of rewriting can
be developed for very tame rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Local
Curch-Rosser is fully recovered for very tame rewriting systems.

For linear rewriting systems sequential independence of derivations
can be also characterised in terms of the existence of a suitable
isomorphism between the colimit of the derivations, the so-called
consistent permutations (which are also strictly related to the notion
of process). This result, which is folklore for categories of graphs,
is explicitly proved here for $\mathcal{M}$-adhesive categories. Based on such result one can deduce two fundamental consequences:
\begin{enumerate}
\item when swicthing a derivation consisting of several steps, the result only depends on the overall permutation, not on the order of the swicthes
\item independence of rewriting steps is a ``global property'', independent of the position in which these steps are in the derivation.
\end{enumerate}

We show that for left-linear rewriting systems the characterisation of
shift equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2). However property (1) can be still proved to holds. Instead only a weak form of (2) can be proved. This is intimately related to the possibility of expressing disnjuctive forms of causality.

>>>>

Rewriting over many kind of graphs is elegantly captured by the
so-called \emph{double pushout} (DPO) approach to rewriting. A
noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their
variants~\cite{lack2005adhesive,ehrig2006weak}. Operating within the
framework of adhesive categories allows one to devise the foundations
of a rewriting theory which can be subsequently instantiated to the
context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results for each specific setting.

Rewriting over adhesive categories has been widely studied in the last
two decades and most of the research focused on the so-called linear
rewriting rules. Roughly speaking, linear rewriting rules can specify
that, in a rewriting step, some entities of the current state are
deleted, some others are needed for the rewriting step to occur but
remain unaltered and some new entities can be generated.

@@

In some situations rewriting with linear rules are not sufficiently
expressive for capturing the dynamics of a system. This is the case
when a computational step is required not only to consume, preserve
and generate, but also ``merge'' parts of the state. This happens,
e.g., in nominal calculi where, as a result of name passing, the
received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanism is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimizations~\cite{WNW:egg}. Rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

@@

Technically, acquiring the expressiveness needed for expressing
fusions, i.e., for merging of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories (which capture most
``applications'') have not been systematically explored. ADD REFS

@@

Firstly, Church-Rosser theorem does not work out of the box. In
particular, in the linear case, sequential independence between
rewriting steps~\cite{} expressed in terms of the existence of a
so-called independence pair is a canonical and well-understood notion
of independence . \todo{P: Too technical? Forse qui e' piu'
  esplicativo un esempio} Whenever two steps are sequential
independent
\begin{itemize}
\item they can be switched and 
\item the result of the switch is uniquely determined.
\end{itemize}

Here we observe that, in general, both properties can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  Since switches are the key notion used to equate computations
which only differs for independent steps, as we will exemplify in the
paper, this phenomenon leads to an inconsistent theory which
identifies clearly different computations.\todo{P: da dire meglio, ma e' un punto centrale}
A refined notion of
independence can be considered to ensure that independent rewriting
steps can be switched, based in the concept of a so-called filler. The
problem was identified already in~\cite{baldan2011adhesivity} where a
suitable subclass of adhesive categories is singled out where
independence as induced by a filler coincide with sequential
independence (as induced by an independence pair). Here this is
extended to $\mathcal{M}$-adhesive categories.

@@

Rather here we identify a class of rewriting systems, the so-called
\emph{very tame rewriting systems} in $\mathcal{M}$-adhesive
categories where sequential independence ensures existence and
uniqueness of the switch and argue that this is the right setting for
dealing with left-linear rewriting systems.

On the one hand we argue that very tame rewriting systems capture most
categories of interest for rewriting. In particular, we single out a
class of very tame left-linear rewriting systems on graph-like
structures sufficiently general to includes many applications of graph
rewriting with ``fusion'' in the literature (roughly, these systems
allows for merging nodes but not edges).  We also show that a class of
rewriting systems on (term)graphs with equivalences resembling
e-graphs are also very tame. As a sanity check, linear rewriting
systems on $\mathcal{M}$-adhesive categories can be shown to be very
tame.

On the other hand we show that an appropriate theory of rewriting can
be developed for very tame rewriting systems, where we recovering,
sometimes in weakened form, results which hold in the linear case. In
particular, local Curch-Rosser theorem is fully recovered for very
tame rewriting systems.

For linear rewriting systems sequential independence of derivations
can be characterised in terms of the existence of a suitable
isomorphism between the colimit of the derivations, the so-called
consistent permutations (strictly related also to the notion of
process). This result, which is known for categories of graphs and
folklore for adhesive categories, is explicitly proved here for linear
rewriting systems on $\mathcal{M}$-adhesive categories. Based on such
result one can deduce two basic properties which are fundamental in the development of the theory of rewriting:
\begin{enumerate}
\item when swicthing a derivation consisting of several steps, the result only depends on the overall permutation, not on the order of the swicthes
\item independence of rewriting steps is a ``global property'', independent of the position in which these steps are in the derivation.
\end{enumerate}

We observe that for left-linear rewriting systems the characterisation of
shift equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2). However, we can show that property (1) still holds, but requires a completely different proof technique. Instead only a weak form of (2) can be proved. This is intimately related to the possibility of expressing disnjuctive forms of causality.
